---
title: "Sample Size + Power"
output: ioslides_presentation
---

```{r setup, include=FALSE}
library(emmeans)
library(tidyverse)
knitr::opts_chunk$set(echo=TRUE)
```

## Purpose of Today

- Understand the expected result of your experiment.
- Are you sampling sufficiently?
- Is your design optimal?

## Type I Error

- Probably of falsely rejecting a TRUE $H_0$.

```{r, echo = FALSE}
  dat.h0 <- data.frame(x = seq(-10,10, by = 0.1))
  dat.h0$f <- dnorm(dat.h0$x, 0, 2 )
plot.pow <- ggplot(dat.h0, aes(x=x, y=f)) + geom_line() +   geom_ribbon(data = dat.h0 %>% filter(x > qnorm(0.975,0,2)), fill = 'green', aes(x=x,ymax=f),ymin=0,alpha=0.3) + geom_vline(aes(xintercept = qnorm(0.975,0,2)), linetype = 'dashed') +
    theme_classic()
plot.pow
```

## Type II Error

- Probably of not rejecting $H_0$, when $H_0$ is false.

```{r, echo = FALSE}
  dat.h1 <- data.frame(x = seq(-5,15, by = 0.1))
  dat.h1$f <- dnorm(dat.h1$x, 4.5, 2 )
plot.pow <- plot.pow + 
    geom_line(data = dat.h1, aes(x=x,y=f), colour = "red") +  
    geom_ribbon(data = dat.h1 %>% filter(x < qnorm(0.975,0,2)), fill = 'red', aes(x=x,ymax=f),ymin=0,alpha=0.3)
plot.pow
```

## Power

- Probability of correctly rejecting a false $H_0$.

```{r, echo = FALSE}
plot.pow <- plot.pow  + 
    geom_ribbon(data = dat.h1 %>% filter(x > qnorm(0.975,0,2)), fill = 'blue', aes(x=x,ymax=f),ymin=0,alpha=0.3)
plot.pow
```

## Controlling Error

- Reduce variance
  - Good design.
  - Pairing observations when possible
  - Consistency in the lab.
- Number of replicates reduces standard error.
- Choosing a reasonable type I Error.
- Large Type I error leads to higher chance of type II error.

## Choosing a sample size

- More samples, smaller standard error.
- We usually have a goal of 80% power.
- Or a level of precision such as (+/- 2.5)
- If we know variance and can guess effect size, we can figure out sample size needed.
- Power is good for an experiment with a hypothesis.
- Precision is better for a field study.

## Power Depends on Effect Size

```{r, echo = FALSE}
dat.h2 <- data.frame(x = seq(-5,15, by = 0.1))
dat.h2$f <- dnorm(dat.h1$x, 6, 2 )

ggplot(dat.h0, aes(x=x, y=f)) + geom_line() + 
  geom_line(data = dat.h1, aes(x=x,y=f), col = 'red') + 
  geom_ribbon(data = dat.h1 %>% filter(x > qnorm(0.975,0,2)), fill = 'red', aes(x=x,ymax=f),ymin=0,alpha=0.3) + geom_vline(aes(xintercept = qnorm(0.975,0,2)), linetype = 'dashed') +
    theme_classic()
```

## Power Depends on Effect Size

```{r echo = FALSE}
ggplot(dat.h0, aes(x=x, y=f)) + geom_line() + 
  geom_line(data = dat.h2, aes(x=x,y=f), col = 'red') + 
  geom_ribbon(data = dat.h2 %>% filter(x > qnorm(0.975,0,2)), fill = 'red', aes(x=x,ymax=f),ymin=0,alpha=0.3) + geom_vline(aes(xintercept = qnorm(0.975,0,2)), linetype = 'dashed') +
    theme_classic()

```

## Power Depends on Variance


```{r, echo = FALSE}
  dat.h0 <- data.frame(x = seq(-15,15, by = 0.1))
  dat.h0$f <- dnorm(dat.h0$x, 0, 4 )

  dat.h1 <- data.frame(x = seq(-5,20, by = 0.1))
  dat.h1$f <- dnorm(dat.h1$x, 6, 4 )

  ggplot(dat.h0, aes(x=x, y=f)) + geom_line() + 
  geom_line(data = dat.h1, aes(x=x,y=f), col = 'red') + 
  geom_ribbon(data = dat.h1 %>% filter(x > qnorm(0.975,0,4)), fill = 'red', aes(x=x,ymax=f),ymin=0,alpha=0.3) + geom_vline(aes(xintercept = qnorm(0.975,0,4)), linetype = 'dashed') +
    theme_classic()
```

## Power of a Z test

- Sample of size $n = 15$, with variance of $\sigma = 3$.
- Null Hypothesis: $\mu_0 = 0$.
- Alternative Hypothesis $\mu_1 > 0$.
- Assume we want to detect a mean of $\mu_1 = 2.5$.
- Significance of $\alpha = 0.05$.

```{r}
  sig.level <- qnorm(0.95, 0, 3/sqrt(15))
  1-pnorm(sig.level, 2.5, 3/sqrt(15))
```

---

```{r}
library(pwr)
pwr.norm.test(d = 2.5/3, n = 15, 
              sig.level = 0.05, alternative = 'greater')
```

## T Test Version

- Things are slightly more complicated as when $H_0$ is false.
- Need to add a non-centrality parameter to the t distribution.

```{r}
  sig.level <- qt(0.95, 14)
  1-pt(sig.level, df = 14, ncp = 2.5/(3/sqrt(15)))
```
---

```{r}
power.t.test(n=15, delta = 2.5,  sd = 3, sig.level = 0.05, 
             type = "one.sample", alternative = "one.sided") 
```

 
## Choosing the Sample Size
 
```{r}
pwr.n <- power.t.test(n=3:50, delta = 2.5,  sd = 3, 
  sig.level = 0.05, type = "one.sample", alternative = "one.sided")
plot(pwr.n$n, pwr.n$power, type = 'l', xlab = "Sample Size", ylab = "Power")
abline(h = 0.8, col = 'red')
```

## Lab Time
 
