---
title: "Randomization + Simulation"
output: html_document
---

## Overview

In this lab we practice how randomize a factorial design. We will simulate an example to show the importance of randomization.

```{r, results= FALSE,message=FALSE}
## Load relevant libraries
library(tidyverse)
#library(sf)
#library(spsurvey)
```

## Experiment Example:

Question: How does competition impact feeding rate of red rock crab?

- Treatment 1: 3 levels of crab density (1 ind, 5 ind, 10 ind).
- Treatment 2: 2 Types of food: (Squid, Mussel)
- Response Variable: Difference in weight of food input before and after 2 hours.
- Design Type: $3\times 2!$ design.
- Replicate 10 times.
- For efficiency, we will block through time in case any variables change.

```{r}
## Define the factorial design.
food <- c("Squid", "Mussel")
density <- c(1, 5, 10)
design <- expand.grid(food = food, density = density, block = 1:10)
```

## Simulation for the results

- Let's assume that the random noise is Normally distributed with standard deviation $\sigma = 0.8$.

```{r}
  params <- data.frame(food = rep(food, each = 3), density = rep(density, 2), mean = c(1.75, 10, 11, 0.9, 1.8, 1.83))
ggplot(data = params, aes(x = density, y = mean, colour = food)) + geom_point()
```

## Randomize the order for each block.

```{r}
design.block.rand <- design %>%
    group_by(block) %>%
    mutate(sampleOrder = sample(n())) %>%
    arrange(sampleOrder, .by_group = TRUE) %>%
    ungroup()
```

## Simulate Some Data

```{r}
data.rand <- design.block.rand %>% 
  group_by(block) %>%
  mutate(block.effect = rnorm(1, 0, 0.2)) %>%
  ungroup() %>%
  left_join(params) %>%
  mutate(response = rnorm(n(), mean + block.effect, 0.8)) %>%
  mutate(density = factor(density))
```

```{r}
ggplot(data = data.rand, aes(x = density, y = response, colour = food)) + geom_boxplot()
```


## Why is the factorial design so important?

Let's first look at this data by estimating the means in each treatment only. In other words, let's run this experiment ignoring interactions and see what the impact of squid is.

```{r}
  mod.add <- aov(response ~ food + density, data = data.rand)
  coef(mod.add)
```

Whoa! When squid is present, waaay more gets eaten. In fact, we expect to see 5.96 g of squid get eaten per crab present compared with mussel. 

## Now let's analyse the data properly.

```{r}
  mod.int <- aov(response ~ food*density, data = data.rand)
  coef(mod.int)
```

Oh, so it turns out that that squid is preferred but it's actually the interaction of increased competition and the better food that results in a much larger consumption by the crabs. The main effect of squid when compared to mussel is only on average 0.57 g, not  5.96 g!!!


## Randomize your own experiment